---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# mlboot

The goal of *mlboot* is to provide a powerful, flexible, and user-friendly way to estimate and compare the performance of machine learning (and other predictive) models using bootstrap resampling. It was created in collaboration by [Jeffrey Girard](https://jmgirard.com/) and [Zhun Liu](http://justin1904.github.io/); this version for R is maintained by Jeffrey Girard and a similar version for Python is maintained by Zhun Liu.

## Installation

``` r
devtools::install_github("jmgirard/mlboot")
```

## Rationale

Bootstrapping is a good choice for estimating the performance of machine learning models because it can be adapted to nearly any type of performance metric, does not make parameteric assumptions about their distributions, and can be quite accurate with relatively little data (e.g., it has been suggested that bootstrapping is appropriate for sample sizes as low as 20, although larger samples obviously confer many benefits). Furthermore, bootstrap confidence intervals provide easily understood information about the precision and reliability of performance estimates and readily extend to statistical comparison.

## Usage

### Estimating the performance of a single model

We can simulate some simple labels and predictions to demonstrate this. Let's say we have data from 1000 videos and we are trying to predict ratings of each video's perceived sentiment (i.e., positivity-versus-negativity) on a scale from 0 to 100. We train a machine learning model on separate data and then generate predictions for each of the 1000 videos just described. We can calculate the performance of this model as the mean absolute error (MAE) across all 1000 videos. However, it would also be nice to know how precise or reliable this estimate is, i.e., how much it is likely to vary as a function of sampling error. To estimate this precision, we can construct a confidence interval around the observed MAE value using bootstrap resampling.

```{r}
# Load the mlboot package
library(mlboot)

# Set random seed for consistent results
set.seed(2020)

# Generate random numbers to simulate trusted labels
ratings <- rnorm(n = 1000, mean = 50, sd = 10)

# Perturb the trusted labels to simulate predictions
model1 <- ratings + rnorm(n = 1000, mean = 10, sd = 10)

# Combine variables into a dataframe
dat <- data.frame(ratings, model1)

# Estimate performance of the simulated predictions using the MAE performance metric
results <- 
  mlboot(
    .data = dat,
    trusted = "ratings",
    predicted = "model1",
    metric = mean_absolute_error
  )

results
```

The output shows that the observed performance (MAE) in the simulated sample was `r round(results$score_obs)`. This is quite close to the perturbation of 10 that we added to create the labels, which is a good sign that our metric function is working properly. The output also shows a 95\% confidence interval around the observed sample statistic; thus, we can be quite confident that the "true" population value of the performance metric is between `r round(results$score_cil, 3)` and `r round(results$score_ciu, 3)`. The p-value of `r round(results$pvalue, 3)` suggests that this MAE scores is significantly different from zero.

### Estimating and comparing the performance of two models
Now let's say we develop another model that is much more accurate. We can use a very similar approach (and indeed the same function call, with additional arguments) to estimate the performance of this second model and assess the degree to which the models differ in performance.

```{r}
# Set random seed for consistent results
set.seed(2020)

# Perturb the trusted labels to a lesser degree to simulate better predictions
model2 <- ratings + rnorm(n = 1000, mean = 8.5, sd = 10)

# Append to existing dataframe
dat2 <- cbind(dat, model2)

# Estimate performance of both models and compare them using the MAE metric
results2 <- 
  mlboot(
    .data = dat2,
    trusted = "ratings",
    predicted = c("model1", "model2"),
    metric = mean_absolute_error,
    pairwise = TRUE
  )

results2
```

The output shows the same observed performance for the first model, although the confidence interval is slightly different due to the stochastic nature of resampling. (If more consistent confidence interval bounds are desired, additional bootstrap resamples can be requested using the `nboot` argument.) The second model had an observed performance score of 5.925 which is indeed lower than that of the first model. To determine whether this difference is statistically significant (i.e., likely to be replicated with another sample), we can estimate the average difference between the performance scores of the models. The observed difference was 4.138 and the confidence interval extends from 3.738 to 4.484 Because the confidence interval does not include zero (and indeed is not particularly close to zero), we can conclude with 95\% confidence that the second model has a lower mean absolute error than the first model.

### Estimating and comparing the performance of many models
The same approach can be applied to any number of models. When `pairwise = TRUE`, all pairs of models will be compared.

```{r}
# Perturb the trusted labels to a lesser degree to simulate better predictions
model3 <- ratings + rnorm(n = 1000, mean = 10, sd = 5)
model4 <- ratings + rnorm(n = 1000, mean = 5, sd = 5)

# Append to existing dataframe
dat3 <- cbind(dat2, model3, model4)

# Estimate performance of both models and compare them using the MAE metric
mlboot(
  .data = dat3,
  trusted = "ratings",
  predicted = c("model1", "model2", "model3", "model4"),
  metric = mean_absolute_error,
  pairwise = TRUE
)
```

### Using the cluster bootstrap for hierarchical data
It is common in many areas of applied machine learning to have testing sets that are hierarchical in structure. For example, there may be multiple testing examples that are clustered (e.g., come from the same individuals or groups) and therefore are not independent. Ignoring this dependency would result in biased estimates, so we need to account for it in some way. Although hierarchical resampling is an active area of research, two recent studies (Field & Welsh, 2007; Ren et al. 2010) suggest that the cluster bootstrap is an accurate and powerful approach to this issue. By supplying a vector indicating cluster membership for each testing example, `mlboot()` can implement the cluster bootstrap procedure. Note that this approach may lead to inaccuracies when the number of clusters is low (e.g., fewer than 20). 

```{r}
# Assume the examples come from 50 different clusters corresponding to persons
person <- rep(1:50, each = 20)

# Generate random numbers to simulate trusted labels
ratings <- rnorm(n = 1000, mean = 20 + person, sd = 10)

# Perturb the trusted labels to simulate predictions
model1 <- ratings + rnorm(n = 1000, mean = 10 - person, sd = 5)
model2 <- ratings + rnorm(n = 1000, mean = 5 - person, sd = 5)

# Combine variables into dataframe
dat3 <- data.frame(person, ratings, model1, model2)

# Estimate and compare the models using the cluster bootstrap
mlboot(
  .data = dat3,
  trusted = "ratings",
  predicted = c("model1", "model2"),
  metric = mean_absolute_error,
  cluster = person,
  pairwise = TRUE
)
```

### Visualizing bootstrap results

```{r histogram}
results <- 
  mlboot(
    .data = dat3,
    trusted = "ratings",
    predicted = c("model1", "model2"),
    metric = mean_absolute_error,
    cluster = person,
    pairwise = TRUE
  )

hist(results$resamples[, 3])
```

## Code of Conduct

Please note that the 'mlboot' project is released with a [Contributor Code of Conduct](.github/CODE_OF_CONDUCT.md). By contributing to this project, you agree to abide by its terms.

## References

Efron, B., & Tibshirani, R. J. (1993). *An introduction to the bootstrap.* New York, NY: Chapman and Hall.

Field, C. A., & Welsh, A. H. (2007). Bootstrapping clustered data. *Journal of the Royal Statistical Society: Series B (Statistical Methodology), 69*(3), 369–390. https://doi.org/10/cqwx5p

Ren, S., Lai, H., Tong, W., Aminzadeh, M., Hou, X., & Lai, S. (2010). Nonparametric bootstrapping for hierarchical data. *Journal of Applied Statistics, 37*(9), 1487–1498. https://doi.org/10/dvfzcn


